---
title: "Identifying weather events with highest impact on health and damages"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

### Synopsis
This report addresses the question, which types of weather events are 1) most harmful with respect to population health and 2) have the greatest economic consequences across the United States. 
To answer these questions, data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database from 1950 to 2011 is used.
    
The data processing consists of removing unnecessary variables, creating two new variables that sum up the consequences to population health and economic consequences and summarizing the effects of weather events grouped by event type. With this processed data, pareto charts containing the TOP 5 weather events for each consequence in question (injuries, fatalities and the sum of these as well as damage to property and crop and the sum of those) are prepared. 

### Data Processing

The analysis is done with data on weather events from 1950 to 2011 from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. 
```{r, message=FALSE, warning=FALSE}
library(lubridate)

# download the data if necessary
if ((!file.exists("StormData.csv")) & (!file.exists("StormData.csv.bz2"))){
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                      "StormData.csv.bz2")
        now()
}

# read in the data
if (!file.exists("StormData.csv")){
        storm_data<-read.csv("StormData.csv.bz2", header=TRUE)
} else {
        storm_data<-read.csv("StormData.csv", header=TRUE)
}

# Dimensions of data set
dim(storm_data)
```

Explanation on the variables used in the downloaded data set can be found here: 
https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

To answer the questions above, only the columns EVTYPE (event type), FATALITIES and INJURIES for the first question and PROPDMG (property damage) and CROPDMG (crop damage) for the second question are needed. All other 32 columns are deleted, the remaining column names are converted to lowercase.

```{r}
# only relevant columns are kept
storm_data<-storm_data[,c(8, 23, 24, 25, 27)]

# for easier use all names are converted to lower
names(storm_data)<-tolower(names(storm_data))
```

To complete the data set, two new variables are added: "health" to sum up the number of fatalities and injuries and "damage" to sum up the costs of property damage and crop damage. The data is then grouped by the event type and the effects to population health and the economic consequences are summarized.
```{r, message=FALSE}
library(dplyr)

# summarizing variables are added
storm_data<-mutate(storm_data, health=fatalities+injuries, damage=propdmg+cropdmg)

# data is grouped by event type and consequences on population health and costs for 
# damages are summarized per event type
data_grouped<-group_by(storm_data,evtype)
sums<-summarize_all(data_grouped, sum)
```

The resulting data set already directly contains the information to answer the questions. 

The following pareto charts containing the TOP 5 weather events per weather consequence in question serve as a basis for a prioritization on which severe weather events to prepare:

```{r}
# creating data sets in decreasing order for each weather consequence in question
# only TOP 5 are used for pareto charts

# population health
health5<-sums[order(sums$health,decreasing=TRUE),][1:5,]
injur5<-sums[order(sums$injuries,decreasing=TRUE),][1:5,]
fatal5<-sums[order(sums$fatalities,decreasing=TRUE),][1:5,]

# costs for damages
dmg5<-sums[order(sums$damage,decreasing=TRUE),][1:5,]
prop5<-sums[order(sums$propdmg,decreasing=TRUE),][1:5,]
crop5<-sums[order(sums$cropdmg,decreasing=TRUE),][1:5,]

# creation of pareto charts as 2 sets of barplots
par(mfrow=c(1,3), las=3, mar=c(9.1, 4.1, 4.1, 2.1), oma=c(2,0,2,0))

# population health
barplot(height=health5$health/1000,names.arg=health5$evtype, 
        ylab="Total of injuries and fatalies caused by event in 1,000", ylim=c(0,100))
title("Injuries and fatalities")

barplot(height=injur5$injuries/1000,names.arg=injur5$evtype, 
        ylab="Total of injuries caused by event in 1,000", ylim=c(0,100))
title("Injuries")

barplot(height=fatal5$fatalities/1000,names.arg=fatal5$evtype, 
        ylab="Total of fatalies caused by event in 1,000", ylim=c(0,100))
title("Fatalities")

title("Top 5 weather events most harmful to population health", outer=TRUE)

# costs for damages
barplot(height=dmg5$damage/1000,names.arg=dmg5$evtype, 
        ylab="Total of property and crop damage costs caused by event in 1,000$", ylim=c(0,4000))
title("Property and crop damage costs")

barplot(height=prop5$propdmg/1000,names.arg=prop5$evtype, 
        ylab="Total of property damage costs caused by event in 1,000$", ylim=c(0,4000))
title("Property damage costs")

barplot(height=crop5$cropdmg/1000,names.arg=crop5$evtype, 
        ylab="Total of crop damage costs caused by event in 1,000$", ylim=c(0,4000))
title("Crop damage costs")

title("Top 5 weather events with greatest economic consequences", outer=TRUE)
```

### Results
The pareto charts show that tornados are - by far - the weather events most harmful to population health. Combining injuries and fatalities, excessive heat and TSTM wind are on second and third place, but effects of tornados are a lot larger.

Looking at injuries only, TSTM wind is on second place and flood takes the third place.

For fatalities excessive heat is on second place, flash floods on third place.

The greatest economic consequences are those which cause the largest costs for property and crop damages. Looking at these costs summed up, tornados again are the event types that have the greatest consequences. Tornados are followed by flash floods and TSTM wind.

Only looking at the costs for property damage, the same order of the leading 3 event types as for the combined economic consequences occurs.

In contrast, for crop damage costs hail is the most harmful weather event. It is followed - with some distance - by flash flood and flood.

